---
title: "Covariance Plot"
output: html_document
editor_options:
  chunk_output_type: console
---

GOAL: Create a plot of the covariance matrix

```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Configuration
```{r}
library(CovStructRM)
library(maditr)
library(ggplot2)
library(assertthat)
library(checkmate)
```

## Example Data Setup
```{r}
data(cholest)

# Convert data between wide and long forms (each subject is now one row)
cholest.w <- dcast(cholest, subjectID + group ~ week, value.var = "ldl")

# Variance-Covariance matrices
cholest.cov <- cov(cholest.w[, 3:ncol(cholest.w)], use = "pairwise.complete.obs", method = "pearson")
```

We need a helper function (`h_get_timepoint_vars`) so that the columns and rows of the input matrix should be named by using the time point values only (ex. "0","3","9","15",...), or by a string (specified in the string argument) separated from the time point value by a period: (ex. "value.0","value.5","value.15",..., "resp.-1","resp.0.5","resp.1","resp.1.5", ...).

```{r}
#' @description get the inputted symmetric matrices row and column labels as numeric time points
#' @param vcov_matrix (`matrix`)\cr name of the input symmetric matrix.
#' @param string (`string`)\cr string in the column/row names of vcov_matrix that precedes the time point value. The default value is NULL,
#' which assumes that the   column/row names of the input matrix don't have any character
#' string other than time point value. If string is specified, this value should appear in the column/row names
#' of the input matrix separated from the time point value by a period.
#' @return This function returns a list with of two sets of numbers: the inputted symmetric matrices row labels and column label time points.
#' @examples
#' h_get_timepoint_vars(vcov_matrix = cholest.cov)
h_get_timepoint_vars <- function(vcov_matrix, string = NULL) {
  assert_string(string, null.ok = TRUE)
  resp_name <- paste(string, ".", sep = "")
  diag <- upper.tri(vcov_matrix, diag = T)
  vect <- vcov_matrix[diag]
  rnames <- rownames(vcov_matrix)
  cnames <- colnames(vcov_matrix)
  nmat <- outer(rnames, cnames, paste, sep = ":")
  names(vect) <- nmat[diag]
  vcov_matrix_row_label <- unlist(strsplit(names(vect), ":"))[seq(1, length(vect) * 2, by = 2)]
  vcov_matrix_col_label <- unlist(strsplit(names(vect), ":"))[seq(2, length(vect) * 2, by = 2)]
  if (is.null(string)) {
    if (any(is.na(as.numeric(vcov_matrix_row_label)))) {
      stop("You have not used the string argument when needed \n
           or the col/row names of your input matrix don't have the time values")
    } else {
      numeric_vcov_matrix_row_label <- as.numeric(vcov_matrix_row_label) # tp.1
      numeric_vcov_matrix_col_label <- as.numeric(vcov_matrix_col_label) # tp.2
    }
  } else {
    if (!any(grepl(string, cnames))) {
      stop("The string you entered is not part of the column/row names of vcov_matrix")
    } else {
      numeric_vcov_matrix_row_label <- as.numeric(gsub(resp_name, "", vcov_matrix_row_label))
      numeric_vcov_matrix_col_label <- as.numeric(gsub(resp_name, "", vcov_matrix_col_label))
    }
  }
  return(list(numeric_vcov_matrix_row_label = numeric_vcov_matrix_row_label, numeric_vcov_matrix_col_label = numeric_vcov_matrix_col_label))
}
```

We need a helper function (`h_vectorization`) for downstream analyses for the plot of the covariance matrix. `h_vectorization` obtains lag and distance between pairs of observations when the column and row names of the input matrix have a numeric value denoting the time point. The columns and rows of the input matrix should be named by using the time point values only (ex. "0","3","9","15",...), or by a string (specified in the string argument) separated from the time point value by a period: (ex. "value.0","value.5","value.15",..., "resp.-1","resp.0.5","resp.1","resp.1.5", ...). The column and row names are obtained from `h_get_timepoint_vars` and so `h_vectorization` just calls this helper function. 

```{r}
#' @description This function vectorizes the upper-diagonal elements of a symmetric matrix (e.g. the covariance matrix)
#' and obtains the lag and time distance between pairs of observations if the time values are part of
#' the names or are part of the matrix column/row names.
#' @param vcov_matrix (`matrix`)\cr name of the input symmetric matrix.
#' @return This function returns a data frame with the upper-diagonal elements or a covariance or correlation
#' matrix. In the context of repeated measures, this matrix contains the association between pairs
#' of measurements taken at different time points.
#' @examples
#' h_vectorization(vcov_matrix = cholest.cov)
h_vectorization <- function(vcov_matrix) {
  assert_matrix(vcov_matrix)
  assert_true(isSymmetric(vcov_matrix))
  diag <- upper.tri(vcov_matrix, diag = T)
  vect <- vcov_matrix[diag]
  timepoints <- h_get_timepoint_vars(vcov_matrix)
  time_point_distribution <- timepoints$numeric_vcov_matrix_col_label - timepoints$numeric_vcov_matrix_row_label
  rank_row <- as.numeric(as.factor(timepoints$numeric_vcov_matrix_row_label))
  rank_col <- as.numeric(as.factor(timepoints$numeric_vcov_matrix_col_label))
  lag <- rank_col - rank_row
  if (any(is.na(lag)) | any(is.na(time_point_distribution))) warning("Verify you have used the string argument correctly or that the columns and rows of your input matrix are named as expected")
  vect <- data.frame(Vect = vect, time_point_distribution, lag, rank_row, rank_col)
  return(vect)
}
```

Now we use this helper function as input for the function to create the covariance plot. 

```{r}
#' @description Plot of covariance or correlation structures as a function of lag or time. The covariance structure
#' needs to be vectorized and lag or time distances computed
#' @param  vcov_dataframe (`data frame`}\cr name of the data frame created by h_vectorization.
#' @param {x_var} (`string`)\cr can be "lag" or "time_point_distribution" for lag and time distance respectively. The default value is "lag".
#' @param legend_position (`string`)\cr denoting where the legend should be shown. Option are c("topright","topleft","bottomright","bottomleft"). Default #' value #' is "topright"
#' @examples
#' g_covariance(vcov_dataframe = h_vectorization(vcov_matrix = cholest.cov), x_var = "time_point_distribution")
g_covariance <- function(vcov_dataframe, x_var = c("lag", "time_point_distribution"), ylab = "",
                         xlab = NULL, col = tre_col, pch = NULL, lty = NULL, cex = 2,
                         legend_position = c("topright", "topleft", "bottomright", "bottomleft"), ...) {
  tre_col <- c(
    "#0080ff", "#ff00ff", "darkgreen", "#ff0000", "orange",
    "#00ff00", "brown"
  )
  x_var <- match.arg(x_var)
  if (x_var == "lag" & is.null(xlab)) {
    xlab <- "Lag"
  }

  if (x_var == "time_point_distribution" & is.null(xlab)) {
    xlab <- "Distance (time units) btw measurements"
  }

  n_col <- length(unique(vcov_dataframe$lag))
  ntp <- 1:max(vcov_dataframe$rank_row)
  if (length(col) < n_col) col <- rep(col, n_col)
  if (is.null(pch)) pch <- ntp
  if (is.null(lty)) lty <- ntp
  legend_position <- match.arg(legend_position)
  vcov_dataframe$rank_row <- as.factor(vcov_dataframe$rank_row)
  ggplot(vcov_dataframe, aes(x = .data[[x_var]], y = .data$Vect, colour = rank_row, group = rank_row)) +
    geom_point() +
    geom_line() +
    labs(colour = "From time:", x = xlab, y = ylab)
}
```

The example obtains covariance plot as expected!
