---
title: "Design for MMRM forest plot"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective

We would like to create a forest plot for the LS means estimated difference
at a specific visit (chosen by the user) across different subgroups created
by different baseline variables. This is FSTG03 "Forest Plot for Adjusted Means"
in the GDSR.

## Details

We will follow the FSTG03 layout from GDSR, with the following
modifications:

- Despite the unclear wording, we see from SAS example code that for each subgroup
  a separate MMRM model is fitted: "Calculate least squares means for each subgroup
  and estimate the mean difference (Treatment minus Control),
  including 95% confidence interval, from an MMRM analysis [...]"
- We replace the column heading "Subgroup" with "Baseline Risk Factor"
- We remove the "Category" column and display the Category headings directly
  underneath the Baseline Risk Factor heading, as in FSTG01.
- Also, we replace the first row heading, "Overall", with "All Patients",
  consistent with FSTG01.
- By default we present the following summary statistics:
  - total N
  - Control n and mean
  - Treatment n and mean
  - mean difference
  - 95% CI for mean difference
- By default plot using constant-sized symbols for the point estimates of the
  differences.  Proportionally sized symbols can be used optionally
  (e.g., size proportional to the total sample size for a subgroup).
- We also allow the option for presenting p-value for each mean difference to the
  right of the 95% CI.
  Note that this is not encouraged, as statistically it does not make
  a lot of sense in most situations - especially since the confidence intervals
  already show whether the p-value is below 5% or not.
- Note that we need exactly one treatment arm and one control arm selected for the
  forest plot. We might have more than one treatment arm in the original dataset
  and model fit. So we need to allow the user to select the treatment arm.

## Idea

We could maybe have a workflow as follows, where we try to stay close
to what we did for the survival and binary response forest plots:

```{r, eval = FALSE}
fit <- fit_mmrm(...)
subgroups <- extract_mmrm_subgroups(
  fit,
  visit = "VIS4",
  subgroups = c("AGEGRP", "REGION", "SEVERITY"),
  groups_lists = list(SEVERITY = list(
    "<= Q1" = c("Q0", "Q1"),
    "> Q1 to <= Q3" = c("Q2", "Q3"),
    "> Q3" = c("Q4", "Q5")
  )),
  label_all = "All Patients"
)
basic_table() %>%
  tabulate_mmrm_subgroups(
    subgroups,
    vars = c("n_tot", "n", "mean", "diff", "ci", "pval")
  )
```

Note:
- The `fit` already contains LS mean estimates, however we will need to derive new
  ones again for the subgroups from separate MMRM model fits. Note that because
  `weights` and `conf_level` are directly taken from the initial overall model fit
  we don't need to fit the overall group again.
- `visit` needs to reference one specific visit.
- We should save he original `cor_struct` in the object so that we can construct
  the subgroup call easily. Same for `parallel` and `accept_singular`.
- We should save the `averages_emmeans` list in the object such that `visit` can
  reference this as a simple string without needing a list specification again.
- We should add the `weights_emmeans` chosen for LS means in the object. However
  it seems like we can only estimate proportional weights. So we might assert that.
- The subgroup variables already need to be factors.
- `groups_lists` helps to group factor levels further as needed (can also overlap).
- The tabulation function should then be relatively similar to the ones for survival
  and binary outcome.
- This must also work when the subgroup variables were not used as part of the
  covariates during the model fit. This should not be a problem now.

# Prototypes

```{r}
library(tern.mmrm)
mmrm_results <- fit_mmrm(
  vars = list(
    response = "FEV1",
    id = "USUBJID",
    arm = "ARMCD",
    visit = "AVISIT"
  ),
  data = mmrm_test_data,
  cor_struct = "unstructured",
  weights_emmeans = "equal",
  averages_emmeans = list(
    "VIS1+2" = c("VIS1", "VIS2")
  )
)
```

## Adjustments to `fit_mmrm`

See the code changes in this pull request.

## LS means computations

This is not an issue anymore as we fit the usual MMRM model just for each subgroup
and then just extract the relevant visit (or average of visits) result for the
given reference and treatment arms.

For the subgroup selection we must also be able to support overlapping subgroups
via the `groups_lists` argument, e.g. for `RACE` we could define (arbitrary example):

```{r}
group_list <- list(
  "A" = c("Asian", "Black or African American"),
  "B" = c("Black or African American", "White")
)
```

## Extraction function

Now that we understand how to calculate the LS means we can write the extraction
function. We can look at the example `?extract_survival_subgroups` to see how
approximately the result should look like to fit into the workflow:

```{r}
# Testing dataset.
library(random.cdisc.data)
library(dplyr)
library(forcats)
library(rtables)

adtte <- cadtte

# Save variable labels before data processing steps.
adtte_labels <- formatters::var_labels(adtte)

adtte_f <- adtte %>%
  filter(
    PARAMCD == "OS",
    ARM %in% c("B: Placebo", "A: Drug X"),
    SEX %in% c("M", "F")
  ) %>%
  mutate(
    # Reorder levels of ARM to display reference arm before treatment arm.
    ARM = droplevels(fct_relevel(ARM, "B: Placebo")),
    SEX = droplevels(SEX),
    AVALU = as.character(AVALU),
    is_event = CNSR == 0
  )
labels <- c(
  "ARM" = adtte_labels[["ARM"]],
  "SEX" = adtte_labels[["SEX"]],
  "AVALU" = adtte_labels[["AVALU"]],
  "is_event" = "Event Flag"
)
formatters::var_labels(adtte_f)[names(labels)] <- labels

df <- extract_survival_subgroups(
  variables = list(
    tte = "AVAL",
    is_event = "is_event",
    arm = "ARM", subgroups = c("SEX", "BMRKR2")
  ),
  data = adtte_f
)
df
```

With that in mind we can prototype the function here.
We need a little helper function for the formatting.

```{r}
h_mmrm_subgroups_df <- function(lsmeans,
                                overall_fit,
                                is_in_subset, # which rows in original data are used?
                                visit,
                                treatment_arm,
                                subgroup,
                                var,
                                label) {
  # Note: If `lsmeans` is `NULL` then no results are available for this subgroup
  # and results are populated with `NA` instead.

  ok <- !is.null(lsmeans)
  v <- overall_fit$vars
  ref_arm <- overall_fit$ref_level

  if (ok) {
    # Estimates.
    estimates <- lsmeans$estimates
    est_selected <- estimates[[v$visit]] == visit &
      estimates[[v$arm]] %in% c(ref_arm, treatment_arm)
    estimates <- estimates[est_selected, ]

    # Contrast.
    contrasts <- lsmeans$contrasts
    cont_selected <- contrasts[[v$visit]] == visit &
      contrasts[[v$arm]] == treatment_arm
    contrasts <- contrasts[cont_selected, ]
  } else {
    subset_model_frame <- stats::model.frame(
      overall_fit$fit$formula_parts$full_formula,
      data = overall_fit$fit$data[is_in_subset, ]
    )
    averages <- attr(overall_fit$lsmeans, "averages")
    visit_selected <- if (visit %in% names(averages)) {
      subset_model_frame[[v$visit]] %in% averages[[visit]]
    } else {
      subset_model_frame[[v$visit]] == visit
    }
    visit_model_frame <- subset_model_frame[visit_selected, ]
    ref_model_frame <- visit_model_frame[visit_model_frame[[v$arm]] == ref_arm, ]
    n_ref <- length(unique(ref_model_frame[[v$id]]))
    trt_model_frame <- visit_model_frame[visit_model_frame[[v$arm]] == treatment_arm, ]
    n_trt <- length(unique(trt_model_frame[[v$id]]))
  }

  # Format as list of two data frames.
  is_all <- var == "ALL"
  row_type <- if (is_all) "content" else "analysis"
  list(
    estimates = data.frame(
      arm = factor(c(ref_arm, treatment_arm), levels = c(ref_arm, treatment_arm)),
      n = if (ok) estimates$n else c(n_ref, n_trt),
      lsmean = if (ok) estimates$estimate else NA,
      subgroup = subgroup,
      var = var,
      var_label = label,
      row_type = row_type
    ),
    contrasts = data.frame(
      n_tot = if (ok) sum(estimates$n) else (n_ref + n_trt),
      diff = if (ok) contrasts$estimate else NA,
      lcl = if (ok) contrasts$lower_cl else NA,
      ucl = if (ok) contrasts$upper_cl else NA,
      pval = if (ok) contrasts$p_value else NA,
      conf_level = overall_fit$conf_level,
      subgroup = subgroup,
      var = var,
      var_label = label,
      row_type = row_type
    )
  )
}
```

Here it finally comes:

```{r}
#' @param fit (`tern_mmrm`)\cr model fit on the total population.
#' @param visit (`string`)\cr single visit or name of averages of visits (referring
#'   to the averages specified when creating the `fit`).
#' @param subgroups (`character` or `NULL`)\cr names of subgroup variables to use in the
#'   forest plot.
#' @param groups_lists (named `list` of `list`)\cr optionally contains for each
#'   `subgroups` variable a list, which specifies the new group levels via the
#'   names and the levels that belong to it in the character vectors that
#'   are elements of the list.
#' @param treatment_arm (`string`)\cr single treatment arm to compare with the reference
#'   arm.ß
#' @param label_all (`string`)\cr label for the total population analysis.
#'
#' @return A list with two elements:
#'   - `estimates`: `data.frame` with columns `arm`, `n`, `lsmean`, `subgroup`,
#'                  `var`, `var_label`, `row_type`.
#'   - `contrasts`: `data.frame` with columns `n_tot`, `diff`, `lcl`, `ucl`,
#'                  `pval`, `subgroup`, `var`, `var_label`, `row_type`. Note
#'                  that this has half the number of rows as `estimates`.
extract_mmrm_subgroups <- function(fit,
                                   visit,
                                   subgroups = NULL,
                                   groups_lists = list(),
                                   treatment_arm = fit$treatment_levels[1L],
                                   label_all = "All Patients") {
  # Assertions to be inserted in production.

  # Shortcuts.
  vars <- fit$vars
  ref_arm <- fit$ref_level
  lsm <- fit$lsmeans
  averages <- attr(lsm, "averages")
  mf <- stats::model.frame(fit$fit, full = TRUE)

  # Get overall result (all patients) - we have it already.
  # Early return if there are no subgroup variables.
  result <- h_mmrm_subgroups_df(
    lsmeans = lsm,
    overall_fit = fit,
    is_in_subset = rep(TRUE, nrow(fit$fit$data)),
    visit = visit,
    treatment_arm = treatment_arm,
    subgroup = label_all,
    var = "ALL",
    label = label_all
  )
  if (is.null(subgroups)) {
    return(result)
  }

  # Now loop over the subgroup variables to calculate subgroup results with
  # separate MMRM fits.
  subgroup_labels <- formatters::var_labels(fit$fit$data[subgroups], fill = TRUE)
  subgroup_results <- lapply(subgroups, function(this_var) {
    checkmate::assert_factor(fit$fit$data[[this_var]])

    this_groups_list <- if (!is.null(groups_lists[[this_var]])) {
      groups_lists[[this_var]]
    } else {
      tmp <- levels(fit$fit$data[[this_var]])
      stats::setNames(as.list(tmp), tmp)
    }

    # Remove this_var from covariates (if present).
    this_vars <- fit$vars
    this_vars$covariates <- setdiff(this_vars$covariates, this_var)

    this_var_results <- list()
    for (this_group in names(this_groups_list)) {
      # Create the data subset.
      this_levels <- this_groups_list[[this_group]]
      dat <- fit$fit$data
      is_in_subset <- dat[[this_var]] %in% this_levels
      this_data <- dat[is_in_subset, , drop = FALSE]

      # Fit the corresponding model.
      this_fit <- tryCatch(
        expr = {
          fit_mmrm(
            vars = this_vars,
            data = this_data,
            cor_struct = fit$cor_struct,
            parallel = fit$parallel,
            accept_singular = fit$accept_singular,
            weights_emmeans = attr(lsm, "weights"),
            averages_emmeans = attr(lsm, "averages")
          )
        },
        error = function(e) {
          message(paste(e, "\n"))
          return(NULL)
        }
      )
      this_var_results[[this_group]] <- h_mmrm_subgroups_df(
        lsmeans = this_fit$lsmeans,
        overall_fit = fit,
        is_in_subset = is_in_subset,
        visit = visit,
        treatment_arm = treatment_arm,
        subgroup = this_group,
        var = this_var,
        label = unname(subgroup_labels[this_var])
      )
    }
    this_var_results
  })
  results <- c(
    list(overall = result),
    unlist(subgroup_results, recursive = FALSE)
  )
  list(
    estimates = do.call(rbind, c(lapply(results, "[[", "estimates"), list(make.row.names = FALSE))),
    contrasts = do.call(rbind, c(lapply(results, "[[", "contrasts"), list(make.row.names = FALSE)))
  )
}
```

So let's try it out:

```{r}
df_mmrm <- extract_mmrm_subgroups(
  fit = mmrm_results,
  visit = "VIS1+2",
  subgroups = c("SEX", "RACE"),
  groups_lists = list(RACE = group_list)
)
df_mmrm
```

That looks good!

Note that we added a try/catch statement for the fact that one
subgroup analysis does not converge. In that case the results should be `NA`
but still be included in the result data frames. Let's try this out too by
adding an artifical factor:

```{r}
dat_fail <- mmrm_test_data
dat_fail$EXTRA <- factor(c(rep("A", 10), rep("B", nrow(dat_fail) - 10)))

# Otherwise same specification here.
mmrm_fail <- fit_mmrm(
  vars = list(
    response = "FEV1",
    id = "USUBJID",
    arm = "ARMCD",
    visit = "AVISIT"
  ),
  data = dat_fail,
  cor_struct = "unstructured",
  weights_emmeans = "equal",
  averages_emmeans = list(
    "VIS1+2" = c("VIS1", "VIS2")
  )
)

# Now the forest function:
df_fail <- extract_mmrm_subgroups(
  fit = mmrm_fail,
  visit = "VIS1+2",
  subgroups = c("SEX", "RACE", "EXTRA"),
  groups_lists = list(RACE = group_list)
)
df_fail
```


## Formatted analysis function

This works very much similarly like `tern::a_survival_subgroups`.

```{r}
a_mmrm_subgroups <- function(.formats) {
  checkmate::assert_list(.formats)
  checkmate::assert_subset(
    names(.formats),
    c("n", "n_tot", "lsmean", "diff", "ci", "pval")
  )

  afun_lst <- Map(
    function(stat, fmt) {
      if (stat == "ci") {
        function(df, labelstr = "", ...) {
          in_rows(
            .list = tern:::combine_vectors(df$lcl, df$ucl),
            .labels = as.character(df$subgroup),
            .formats = fmt
          )
        }
      } else {
        function(df, labelstr = "", ...) {
          in_rows(
            .list = as.list(df[[stat]]),
            .labels = as.character(df$subgroup),
            .formats = fmt
          )
        }
      }
    },
    stat = names(.formats),
    fmt = .formats
  )

  afun_lst
}
```

## Table creating function

This works similarly as `tern::tabulate_survival_subgroups`.

```{r}
tabulate_mmrm_subgroups <- function(lyt,
                                    df,
                                    vars = c("n_tot", "n", "lsmean", "diff", "ci"),
                                    .formats = list(
                                      n = "xx",
                                      n_tot = "xx",
                                      lsmean = "xx.x",
                                      diff = "xx.x",
                                      ci = "(xx.x, xx.x)",
                                      pval = "x.xxxx | (<0.0001)"
                                    ),
                                    .labels = list(
                                      n = "n",
                                      n_tot = "Total n",
                                      lsmean = "Mean",
                                      diff = "Mean Difference",
                                      ci = paste0(round(100 * df$contrasts$conf_level[1]), "% CI"),
                                      pval = "p-value"
                                    )) {
  afun_lst <- a_mmrm_subgroups(.formats = .formats)
  var_labels <- unlist(.labels)
  checkmate::assert_subset(x = c("n_tot", "diff", "ci"), choices = vars)

  colvars <- vars
  # The `lcl` variable is just a placeholder available in the analysis data,
  # it is not acutally used in the tabulation.
  # Variables used in the tabulation are lcl and ucl, see `a_mmrm_subgroups` for details.
  colvars[colvars == "ci"] <- "lcl"
  colvars <- list(
    vars = colvars,
    labels = var_labels[vars]
  )
  is_estimates <- names(colvars$labels) %in% c("n", "lsmean")
  colvars_estimates <- list(
    vars = colvars$vars[is_estimates],
    labels = colvars$labels[is_estimates]
  )
  is_contrasts <- names(colvars$labels) %in% c("n_tot", "diff", "ci", "pval")
  colvars_contrasts <- list(
    vars = colvars$vars[is_contrasts],
    labels = colvars$labels[is_contrasts]
  )

  # Columns from table_survtime are optional.
  if (length(colvars_estimates$vars) > 0) {
    lyt_estimates <- split_cols_by(lyt = lyt, var = "arm")
    lyt_estimates <- split_rows_by(
      lyt = lyt_estimates,
      var = "row_type",
      split_fun = keep_split_levels("content"),
      nested = FALSE
    )
    lyt_estimates <- summarize_row_groups(
      lyt = lyt_estimates,
      var = "var_label",
      cfun = afun_lst[names(colvars_estimates$labels)]
    )
    lyt_estimates <- split_cols_by_multivar(
      lyt = lyt_estimates,
      vars = colvars_estimates$vars,
      varlabels = colvars_estimates$labels
    )

    if ("analysis" %in% df$estimates$row_type) {
      lyt_estimates <- split_rows_by(
        lyt = lyt_estimates,
        var = "row_type",
        split_fun = keep_split_levels("analysis"),
        nested = FALSE,
        child_labels = "hidden"
      )
      lyt_estimates <- split_rows_by(lyt = lyt_estimates, var = "var_label", nested = TRUE)
      lyt_estimates <- analyze_colvars(
        lyt = lyt_estimates,
        afun = afun_lst[names(colvars_estimates$labels)],
        inclNAs = TRUE
      )
    }

    table_estimates <- build_table(lyt_estimates, df = df$estimates)
  } else {
    table_estimates <- NULL
  }

  # Columns "n_tot", "diff", "ci" in table_contrasts are required.
  # lyt_contrasts <- split_cols_by(lyt = lyt, var = "arm")
  lyt_contrasts <- split_rows_by(
    lyt = lyt,
    var = "row_type",
    split_fun = keep_split_levels("content"),
    nested = FALSE
  )
  lyt_contrasts <- summarize_row_groups(
    lyt = lyt_contrasts,
    var = "var_label",
    cfun = afun_lst[names(colvars_contrasts$labels)]
  )
  lyt_contrasts <- split_cols_by_multivar(
    lyt = lyt_contrasts,
    vars = colvars_contrasts$vars,
    varlabels = colvars_contrasts$labels
  ) %>%
    append_topleft("Baseline Risk Factors")

  if ("analysis" %in% df$contrasts$row_type) {
    lyt_contrasts <- split_rows_by(
      lyt = lyt_contrasts,
      var = "row_type",
      split_fun = keep_split_levels("analysis"),
      nested = FALSE,
      child_labels = "hidden"
    )
    lyt_contrasts <- split_rows_by(lyt = lyt_contrasts, var = "var_label", nested = TRUE)
    lyt_contrasts <- analyze_colvars(
      lyt = lyt_contrasts,
      afun = afun_lst[names(colvars_contrasts$labels)],
      inclNAs = TRUE
    )
  }
  table_contrasts <- build_table(lyt_contrasts, df = df$contrasts)

  n_tot_id <- grep("n_tot", colvars_contrasts$vars, fixed = TRUE)
  if (is.null(table_estimates)) {
    result <- table_contrasts
    diff_id <- match("diff", colvars_contrasts$vars)
    ci_id <- match("lcl", colvars_contrasts$vars)
  } else {
    # Reorder the table.
    result <- cbind_rtables(table_contrasts[, n_tot_id], table_estimates, table_contrasts[, -n_tot_id])
    # And then calculate column indices accordingly.
    diff_id <- 1L + ncol(table_estimates) + match("diff", colvars_contrasts$vars[-n_tot_id])
    ci_id <- 1L + ncol(table_estimates) + match("lcl", colvars_contrasts$vars[-n_tot_id])
    n_tot_id <- 1L
  }

  structure(
    result,
    forest_header = paste0(levels(df$estimates$arm), "\nHigher"),
    col_x = diff_id,
    col_ci = ci_id,
    col_symbol_size = n_tot_id
  )
}
```

Let's try this out.

```{r}
tab_mmrm <- basic_table() %>%
  tabulate_mmrm_subgroups(df = df_mmrm)
```

Then we can plot it too.

```{r}
g_forest(tab_mmrm, logx = FALSE, xlim = c(-10, 10), x_at = c(-10, -5, 0, 5, 10), vline = 0)
```

To do: In production we can make a small wrapper `g_forest_mmrm` which sets
meaningful defaults for the arguments of `g_forest`.
