---
title: "Design for MMRM forest plot"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective

We would like to create a forest plot for the LS means estimated difference
at a specific visit (chosen by the user) across different subgroups created
by different baseline variables. This is FSTG03 "Forest Plot for Adjusted Means"
in the GDSR.

## Details

We will follow the FSTG01 layout example from GDSR, with the following
modifications:

- We note that only one MMRM is fitted and then the LS means are extracted
  from this single MMRM fit: "Calculate least squares means for each subgroup
  and estimate the mean difference (Treatment minus Control),
  including 95% confidence interval, from an MMRM analysis [...]"
- We replace the column heading "Subgroup" with "Baseline Risk Factor"
- We remove the "Category" column and display the Category headings directly
  underneath the Baseline Risk Factor heading, as in FSTG01.
- Also, we replace the first row heading, "Overall", with "All Patients",
  consistent with FSTG01.
- By default we present the following summary statistics:
  - total N
  - Control n and mean
  - Treatment n and mean
  - mean difference
  - 95% CI for mean difference
- By default plot using constant-sized symbols for the point estimates of the
  differences.  Proportionally sized symbols can be used optionally
  (e.g., size proportional to the total sample size for a subgroup).
- We also allow the option for presenting p-value for each mean difference to the
  right of the 95% CI.
  Note that this is not encouraged, as statistically it does not make
  a lot of sense in most situations - especially since the confidence intervals
  already show whether the p-value is below 5% or not.

## Idea

We could maybe have a workflow as follows, where we try to stay close
to what we did for the survival and binary response forest plots:

```{r, eval = FALSE}
fit <- fit_mmrm(...)
subgroups <- extract_mmrm_subgroups(
  fit,
  visit = "VIS4",
  subgroups = c("AGEGRP", "REGION", "SEVERITY"),
  groups_lists = list(SEVERITY = list(
    "<= Q1" = c("Q0", "Q1"),
    "> Q1 to <= Q3" = c("Q2", "Q3"),
    "> Q3" = c("Q4", "Q5")
  )),
  weights = fit$weights,
  conf_level = fit$conf_level,
  label_all = "All Patients"
)
basic_table() %>%
  tabulate_mmrm_subgroups(
    subgroups,
    vars = c("n_tot", "n", "mean", "diff", "ci", "pval")
  )
```

Note:
- The `fit` already contains LS mean estimates, however we will need to derive new
  ones again for the subgroups.
- `visit` needs to reference one specific visit.
- We should save the `averages_emmeans` list in the object such that `visit` can
  reference this as a simple string without needing a list specification again.
- We should add the `weights_emmeans` chosen for LS means in the object.
- The subgroup variables already need to be factors.
- `groups_lists` helps to group factor levels further as needed (can also overlap).
- The tabulation function should then be relatively similar to the ones for survival
  and binary outcome.

# Prototypes

## Adjustments to `fit_mmrm`

See the code changes in this pull request.

## Extraction function

We will need to figure out here how to calculate the LS means for a specified
subgroup.
One additional challenge is when an average of visits is requested instead of
a single visit.

## Formatted analysis function

## Table creating function
