---
title: "Satterthwaite DF Example"
output: html_document
editor_options:
  chunk_output_type: console
---

GOAL: To see if we get similar results between `lme4`, `glmmTMB` and `SAS`
(not considering the df correction that needs to be done).

```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Configuration

```{r setup, message=FALSE, warnings = FALSE}
library(rtables)
library(tern.mmrm)
library(dplyr)
library(r2stream)
library(scda)
library(glmmTMB)
library(emmeans)
```

## Data for example comparison


```{r}
adqs <- synthetic_cdisc_data("latest")$adqs
adqs_f <- adqs %>%
  dplyr::filter(PARAMCD == "FKSI-FWB" & AVISIT %in% c("WEEK 1 DAY 8", "WEEK 2 DAY 15", "WEEK 3 DAY 22")) %>%
  droplevels() %>%
  dplyr::mutate(ARMCD = factor(ARMCD, levels = c("ARM B", "ARM A", "ARM C"))) %>%
  dplyr::mutate(
    AVISITN = rank(AVISITN) %>%
      as.factor() %>%
      as.numeric() %>%
      as.factor()
  )
formatters::var_labels(adqs_f) <- formatters::var_labels(adqs)
```


## Model fitting

Results: Model did not converge for both `scda` and `random.cdisc.data` using all `AVISIT` values in datasets. Hence, for this example comparison, only "WEEK 1 DAY 8",  "WEEK 2 DAY 15", "WEEK 3 DAY 22" were filtered in the dataset.

### Using the lme4 / lmerTest package to get the Satterthwaite df (Current `tern.mmrm` implementation)

#### Least Squares Means

```{r}
mmrm_results <- fit_mmrm(
  vars = list(
    response = "AVAL",
    covariates = c("STRATA1", "BMRKR2"),
    id = "USUBJID",
    arm = "ARMCD",
    visit = "AVISIT"
  ),
  data = adqs_f,
  cor_struct = "unstructured",
  weights_emmeans = "equal"
)

df <- tidy(mmrm_results)
attr(df$AVISIT, "label") <- "Visit"

# Define the split function
split_fun <- drop_split_levels

result <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = mmrm_results$ref_level) %>%
  add_colcounts() %>%
  split_rows_by("AVISIT", split_fun = split_fun, label_pos = "topleft", split_label = obj_label(df$AVISIT)) %>%
  summarize_lsmeans(show_relative = "increase") %>%
  append_topleft("  Statistics") %>%
  build_table(df)

result

lsmeans_lme4 <- mmrm_results$lsmeans$estimate
```

#### Coefficient Estimates

```{r}
summary(mmrm_results$fit)
```

### Using SAS

```{r, results='hide', eval = FALSE}
sascode <- list(
  # "show" = "
  #   PROC CONTENTS DATA = ana.dat;
  #   RUN;
  # "
  "test" = "
    PROC MIXED DATA = ana.dat cl method=reml;
      CLASS USUBJID STRATA1(ref = 'A') BMRKR2(ref='LOW') AVISIT(ref = 'WEEK 1 DAY 8') ARMCD(ref = 'ARM B');
      MODEL AVAL = STRATA1 BMRKR2 ARMCD AVISIT ARMCD*AVISIT / ddfm=satterthwaite solution chisq;
      REPEATED AVISIT / subject=USUBJID type=un r rcorr;
      LSMEANS ARMCD*AVISIT / pdiff=all cl alpha=0.05 slice=AVISIT;
    RUN;
      "
)


result <- r2stream::bee_sas(data = list("dat" = adqs_f), sascode = sascode)
```

```{r, eval = FALSE}
result$test$sas_log
result$test$sas_out

# ""
# [588] "                                                        Least Squares Means"
# [589] " "
# [590] "                                  Planned"
# [591] "                 Analysis         Arm                    Standard"
# [592] " Effect          Visit            Code       Estimate       Error      DF    t Value    Pr > |t|     Alpha       Lower       Upper"
# [593] ""
# [594] " AVISIT*ARMCD    WEEK 2 DAY 15    ARM A       59.8226      0.7622     391      78.49      <.0001      0.05     58.3241     61.3210"
# [595] " AVISIT*ARMCD    WEEK 2 DAY 15    ARM C       59.3272      0.7674     390      77.31      <.0001      0.05     57.8184     60.8361"
# [596] " AVISIT*ARMCD    WEEK 2 DAY 15    ARM B       59.8945      0.7629     392      78.50      <.0001      0.05     58.3945     61.3944"
# [597] " AVISIT*ARMCD    WEEK 3 DAY 22    ARM A       66.8182      0.8741     393      76.44      <.0001      0.05     65.0996     68.5367"
# [598] " AVISIT*ARMCD    WEEK 3 DAY 22    ARM C       66.6071      0.8803     392      75.66      <.0001      0.05     64.8764     68.3378"
# [599] " AVISIT*ARMCD    WEEK 3 DAY 22    ARM B       64.3064      0.8748     394      73.51      <.0001      0.05     62.5865     66.0262"
# [600] " AVISIT*ARMCD    WEEK 1 DAY 8     ARM A       55.0505      0.6896     393      79.83      <.0001      0.05     53.6948     56.4062"
# [601] " AVISIT*ARMCD    WEEK 1 DAY 8     ARM C       53.9807      0.6942     392      77.76      <.0001      0.05     52.6158     55.3456"
# [602] " AVISIT*ARMCD    WEEK 1 DAY 8     ARM B       54.2473      0.6904     394      78.57      <.0001      0.05     52.8899     55.6046"
# [603] ""
# [604] ""
# [605] "                                                Differences of Least Squares Means"
# [606] " "
# [607] "                                   Planned                   Planned"
# [608] "                   Analysis        Arm       Analysis        Arm                  Standard"
# [609] "    Effect         Visit           Code      Visit           Code      Estimate      Error     DF   t Value   Pr > |t|    Alpha"
# [610] ""
# [611] "    AVISIT*ARMCD   WEEK 2 DAY 15   ARM A     WEEK 2 DAY 15   ARM C       0.4953     1.0810    390      0.46     0.6471     0.05"
# [612] "    AVISIT*ARMCD   WEEK 2 DAY 15   ARM A     WEEK 2 DAY 15   ARM B     -0.07192     1.0795    393     -0.07     0.9469     0.05"
# [613] "    AVISIT*ARMCD   WEEK 2 DAY 15   ARM A     WEEK 3 DAY 22   ARM A      -6.9956     1.1478    397     -6.09     <.0001     0.05"
# [614] "    AVISIT*ARMCD   WEEK 2 DAY 15   ARM A     WEEK 3 DAY 22   ARM C      -6.7846     1.1639    772     -5.83     <.0001     0.05"



#  "                                                     Solution for Fixed Effects"
# [537] " "
# [538] "                                    Categorical                     Planned"
# [539] "                  Stratification    Level          Analysis         Arm                    Standard"
# [540] "  Effect          Factor 1          Biomarker 2    Visit            Code       Estimate       Error      DF    t Value    Pr > |t|"
# [541] ""
# [542] "  Intercept                                                                     54.3982      0.8644     536      62.93      <.0001"
# [543] "  STRATA1         B                                                              0.2776      0.6463     393       0.43      0.6678"
# [544] "  STRATA1         C                                                             0.04540      0.6372     393       0.07      0.9432"
# [545] "  STRATA1         A                                                                   0           .       .        .         .    "
# [546] "  BMRKR2                            HIGH                                        -0.9818      0.6370     393      -1.54      0.1241"
# [547] "  BMRKR2                            MEDIUM                                       0.2061      0.6324     393       0.33      0.7447"
# [548] "  BMRKR2                            LOW                                               0           .       .        .         .    "
# [549] "  ARMCD                                                             ARM A        0.8033      0.9770     395       0.82      0.4115"
# [550] "  ARMCD                                                             ARM C       -0.2666      0.9801     394      -0.27      0.7858"
# [551] "  ARMCD                                                             ARM B             0           .       .        .         .    "
# [552] "  AVISIT                                           WEEK 2 DAY 15                 5.6472      1.0331     397       5.47      <.0001"
# [553] "  AVISIT                                           WEEK 3 DAY 22                10.0591      1.0897     397       9.23      <.0001"
# [554] "  AVISIT                                           WEEK 1 DAY 8                       0           .       .        .         .    "
# [556] ""
# [557] "                                                        The Mixed Procedure"
# [558] ""
# [559] "                                                     Solution for Fixed Effects"
# [560] " "
# [561] "                                    Categorical                     Planned"
# [562] "                  Stratification    Level          Analysis         Arm                    Standard"
# [563] "  Effect          Factor 1          Biomarker 2    Visit            Code       Estimate       Error      DF    t Value    Pr > |t|"
# [564] ""
# [565] "  AVISIT*ARMCD                                     WEEK 2 DAY 15    ARM A       -0.8752      1.4610     397      -0.60      0.5495"
# [566] "  AVISIT*ARMCD                                     WEEK 2 DAY 15    ARM C       -0.3007      1.4665     397      -0.21      0.8377"
# [567] "  AVISIT*ARMCD                                     WEEK 2 DAY 15    ARM B             0           .       .        .         .    "
# [568] "  AVISIT*ARMCD                                     WEEK 3 DAY 22    ARM A        1.7085      1.5411     397       1.11      0.2682"
# [569] "  AVISIT*ARMCD                                     WEEK 3 DAY 22    ARM C        2.5673      1.5469     397       1.66      0.0978"
# [570] "  AVISIT*ARMCD                                     WEEK 3 DAY 22    ARM B             0           .       .        .         .    "
# [571] "  AVISIT*ARMCD                                     WEEK 1 DAY 8     ARM A             0           .       .        .         .    "
# [572] "  AVISIT*ARMCD                                     WEEK 1 DAY 8     ARM C             0           .       .        .         .    "
# [573] "  AVISIT*ARMCD                                     WEEK 1 DAY 8     ARM B             0           .       .        .         .    "
```




### Using glmmTMB

#### Least Square Means

```{r}
fit_glmmtmb <- glmmTMB(
  formula = AVAL ~ STRATA1 + BMRKR2 + ARMCD + ARMCD + AVISIT + ARMCD * AVISIT + us(0 + AVISIT | USUBJID),
  data = adqs_f,
  dispformula = ~0,
  REML = TRUE, # whether to use REML estimation rather than maximum likelihood.
  # start = NULL,
  control = glmmTMB::glmmTMBControl(optimizer = optim, optArgs = list(method = "L-BFGS-B"), parallel = 1)
)

lsmeans_glmmtmb <- emmeans(fit_glmmtmb,
  specs = ~ ARMCD | AVISIT,
  data = adqs_f
)
lsmeans_glmmtmb
```

We can also look at the contrasts then easily:

```{r}
pairs(lsmeans_glmmtmb)
```

(Note that here of course the degrees of freedom are wrong.)

#### Coefficient Estimates

These can be obtained with:

```{r}
summary(fit_glmmtmb)
```

### Visit averages with emmeans

Let's try now to calculate averages of visits with `emmeans` for the
`glmmTMB` result.

The `contrast()` function in `emmeans` can be used to compute any linear function
of estimated marginal means (EMM), see https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html#linfcns
for details.

#### Starting point

For this we cannot use the formula interface for `emmeans` as above, but first
just specify the two variables involved here, i.e. arm and visit:

```{r}
lsmeans_general <- emmeans(fit_glmmtmb,
  c("ARMCD", "AVISIT"),
  data = adqs_f
)
lsmeans_general
```

#### Specifying coefficients for grid

Now based on this result we see the order of the marginal means, and similarly
as in SAS, can specify the required linear combinations across the grid of treatment
arms and visits:

```{r}
average_specs <- list(
  "ARM B Average W1D8 and W2D15" = c(1, 0, 0, 1, 0, 0, 0, 0, 0) / 2,
  "ARM A Average W1D8 and W2D15" = c(0, 1, 0, 0, 1, 0, 0, 0, 0) / 2,
  "ARM C Average W1D8 and W2D15" = c(0, 0, 1, 0, 0, 1, 0, 0, 0) / 2
)
contrast(
  lsmeans_general,
  average_specs
)
```

Note that we can see the grid by looking at the corresponding slot of the S4
object:

```{r}
lsmeans_general@grid
```

We can also derive this list a bit more automatically.

```{r}
visit_grid_levels <- lsmeans_general@grid$AVISIT
visits_average_specs <- list(
  "Average W1D8, W2D15" = c("WEEK 1 DAY 8", "WEEK 2 DAY 15"),
  "Average W2D15, W3D22" = c("WEEK 2 DAY 15", "WEEK 3 DAY 22")
)
average_grid_list <- list()
for (i_spec in seq_along(visits_average_specs)) {
  average_label <- names(visits_average_specs)[i_spec]
  visits_average <- visits_average_specs[[i_spec]]
  average_coefs <- as.integer(visit_grid_levels %in% visits_average) / length(visits_average)
  zero_coefs <- numeric(length = length(average_coefs))
  for (this_arm in levels(lsmeans_general@grid$ARMCD)) {
    arm_average_label <- paste(this_arm, average_label)
    this_coefs <- zero_coefs
    this_index <- which(lsmeans_general@grid$ARMCD == this_arm)
    this_coefs[this_index] <- average_coefs[this_index]
    average_grid_list[[arm_average_label]] <- this_coefs
  }
}
average_conts <- contrast(
  lsmeans_general,
  average_grid_list
)
```

Obviously with this we can also estimate differences between the treatment arms.
For example, comparing both arms A and C with arm B (as that is labelled "Placebo"):

```{r}
diff_specs <- list(
  "ARM A vs. ARM B for average W1D8 and W2D15" = average_specs[[2]] - average_specs[[1]],
  "ARM C vs. ARM B for average W1D8 and W2D15" = average_specs[[3]] - average_specs[[1]]
)
contrast(
  lsmeans_general,
  diff_specs
)
```

And actually it is even easier: We can call `pairs()` here.

```{r}
pairs(average_conts)
```

Only problem is that we have too many contrasts then here again. So probably
specifying them manually is more efficient while needing a bit more code first.

#### Coefficients for visits by arm

It is also possible to specify a `by` variable here which can be useful
potentially:

```{r}
contrast(
  lsmeans_general,
  list(
    "Average W1D8 and W2D15" = c(1, 1, 0) / 2
  ),
  by = "ARMCD"
)
```

We could derive this a bit more automatically like this:

```{r}
visit_levels <- lsmeans_general@levels$AVISIT
visits_average <- c("WEEK 1 DAY 8", "WEEK 2 DAY 15")
average_label <- paste("Average", toString(visits_average))
average_coefs <- as.integer(visit_levels %in% visits_average) / length(visits_average)
average_list <- setNames(list(average_coefs), average_label)

cont_by_arm <- contrast(
  lsmeans_general,
  average_list,
  by = "ARMCD"
)
cont_by_arm
```

The question is whether we can get differences now?

```{r}
str(cont_by_arm)
class(cont_by_arm)
# pairs(cont_by_arm)
# does not work unfortunately

pairs(lsmeans_general, by = "AVISIT")
```

#### New grouped factor

Another helpful tool could be to use grouping of factor levels.

```{r}
levels(lsmeans_general)
lsmeans_with_grp <- add_grouping(
  lsmeans_general,
  newname = "AVISITGRP",
  refname = "AVISIT",
  newlevs = c("AVG W1D8 and W2D15", "AVG W1D8 and W2D15", NA)
)
lsmeans_by_avg_visit <- emmeans(lsmeans_with_grp, ~ ARMCD | AVISITGRP)
lsmeans_by_avg_visit
pairs(lsmeans_by_avg_visit)
```

So it is reassuring that the results are the same as above here.
It is just a bit inconvenient that we then only get the results for this new
average, rather than also for the other visits.


### Summary of Results

In conclusion, the results between `lme4`,`glmmTMB` and `SAS` are pretty much identical up to at least 2 decimal places for the least square means, standard deviation, and confidence intervals. The model coefficient estimates are the same between `lme4`, `glmmTMB` and `SAS` to two and in some cases even to 3 decimal places.
